Jane, John … Leslie? A Historical Method for Algorithmic Gender Detection

### 2014-09-07 article.Rmd
Description of project history

The most significant problem with the Kantrowitz names corpus and thus the NLTK implementation is that it assumes that names are timeless. As pointed out above, this makes the corpus unusable for historical purposes. Furthermore the corpus includes some oddities which make it less than suitable. Some names like Abby are overwhelmingly female and some like Bill are overwhelmingly male, but the corpus includes them as both male and female. The Kantrowitz corpus contains only `r kantrowitz_unique`, a mere `r percent(kantrowitz_unique / ssa_unique)` of the `r ssa_unique` names in the Social Security Administration dataset and `r percent(kantrowitz_unique / ipums_unique)` of the `r ipums_unique` names in the census records provided by IPUMS USA. There are therefore many names that it cannot identify. Assuming for the moment that our method provides more accurate results, we estimate that `r percent(kantrowitz_error_rate)` percent of the names in the Kantrowitz corpuse are misclassified or classified as ambigious when a prediction could be made.

@@

To that end we have created the [Gender](https://github.com/ropensci/gender) package for R which contains the , and an associated [genderdata](https://github.com/lmullen/gender-data-pkg) package which contains various historical datasets.

The possibilities for encoding gender from names depends on two things. First, it depends on having a suitable (and suitably large) dataset for the time period and region under study. Unsurprisingly such datasets are almost always gathered in the first instance by states, though the compilation and digitization may be undertaken by scholarly researchers. Second, it depends on a suitable algorithm for estimating the proportion of male and female names for a given year, or range of years since often a person cannot be associated with an exact date. It is especially important that the algorithm take into account any biases in the data to make correct predictions.

Besides including the

>>>>

The most significant problem with the Kantrowitz names corpus and thus the NLTK implementation is that it assumes that names are timeless. As pointed out above, this makes the corpus unusable for historical purposes. Furthermore the corpus includes some oddities which make it less than suitable. Some names like Abby are overwhelmingly female and some like Bill are overwhelmingly male, but the corpus includes them as both male and female. The Kantrowitz corpus contains only `r kantrowitz_unique`, a mere `r percent(kantrowitz_unique / ssa_unique)` of the `r ssa_unique` names in the Social Security Administration dataset and `r percent(kantrowitz_unique / ipums_unique)` of the `r ipums_unique` names in the census records provided by IPUMS USA. There are therefore many names that it cannot identify. Assuming for the moment that our method provides more accurate results, we estimate that `r percent(kantrowitz_error_rate)` percent of the names in the Kantrowitz corpuse are misclassified or classified as ambigious when a valid prediction could be made.

@@

To that end we have created the [Gender](https://github.com/ropensci/gender) package for R which contains the predictive algorithm, and an associated [genderdata](https://github.com/lmullen/gender-data-pkg) package which contains various historical datasets. This R implementation is based on an [earlier Python implementation](https://github.com/cblevins/Gender-ID-By-Time) by [Cameron Blevins](http://www.cameronblevins.org/) and [Bridget Baird](http://www.conncoll.edu/directories/emeritus-faculty/bridget-baird/).

The possibilities for encoding gender from names depends on two things. First, it depends on having a suitable (and suitably large) dataset for the time period and region under study. Unsurprisingly such datasets are almost always gathered in the first instance by states, though the compilation and digitization may be undertaken by scholarly researchers. Second, it depends on a suitable algorithm for estimating the proportion of male and female names for a given year, or range of years since often a person cannot be associated with an exact date. It is especially important that the algorithm take into account any biases in the data to make correct predictions. Development of the R package has had two primary aims. The first is to abstract the predictive algorithm to the simplest possible form so that it is usable for a wide range of historical problems rather than depending on the format of any particular data set. The second has been to provide as many datasets as possible in order to localize the predictions to particular times and places. To that end, [Ben Schmidt](http://benschmidt.org/) contributed a dataset of US Census Data and accompanying code, as well as providing a significant correction to predictions based on Social Security Administration data.
